Tidy data structure to support exploration and modeling of temporal-context data

### 2018-07-03 main.Rmd
added a bit more on how to arrange the toy example

In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to different representations. A matrix is used to represent multivariate time series where each row represents observations measured at a time point and each column represents a series ("wide" form). This matrix representation requires homogeneity (that is, all the columns must be of same type.), time indices implicitly inferred as attributes or meta-information, series of same time length, and explicit missingness. By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm** [@R-plm]. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. [R]

@@

# Data semantics

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represent time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" works similarly as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. A syntax is introduced to express a key consisting of nested and crossed factors. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database.

Given the nature of temporal ordering, a temporal data set must be sorted by time index. If a key is explicitly declared, the key will be sorted first and followed by arranging time in ascending order.

@@

```{r load-data}
```

(ref:tb-sub) A subset of estimates of tuberculosis burden generated by World Health Organisation in 2011 and 2012.

## Time index and interval

@@

For data indexed in regular time space, the time interval is obtained by computing the greatest common divisor based on the non-negative differences of a given time index, although the data may include implicit time gaps. This implies that all the units in the table must share a common interval. It is relatively easy to determine if the data is daily or sub-daily since most statistical packages have built-in time representations for date-times and dates, for example their corresponding classes `POSIXct` and `Date` in R. However these turn out inappropriate for weekly, monthly and quarterly data. For example, the smallest time unit of monthly data is months in a year, which is equispaced points in a year. However, using the first day of the month can range from 28 to 31 days, which is irregularly spaced. A genuine year-month representation should avoid associating days with its format. Displaying the year and month components suffice to signal its monthly aggregated values instead of observed values at a particular timestamp.

@@

## Key

The "key" identifies units or subjects that are observed over time in a data table, which are typically known in advance by users. The absence of key only occurs to a univariate case where the key is implicit. Multiple units must associate with an explicit "key" to allow for identification. The key is not constrained to a single column, but is comprised of multiple columns. When involving multiple variables, one need to distinguish nesting and crossing data variables. "Nesting" refers to a scenario where a variable is nested within another when each category of the former variable co-occurs with only one category of the latter; on the other hand, "crossing" means that a variable is crossed with another when every category of one variable co-occurs with every category of the other. Nesting is a special case of crossing. The former exhibits a hierarchical ordering with lower-level category (children) nested within higher-level category (parent); whereas the ordering of the latter makes no difference. Whilst constructing tidy temporal data, users also need to specify the key that identifies the structure on the data. Rather than creating a separate object that contains the data structure, like what the **hts** does, we propose a syntax-based interface to directly deal with variables, so that it takes advantage of data semantics. The expression of a vertical bar (`|`) is used to indicate nesting variables, while a comma (`,`) for crossing variables. @wilkinson2006grammar discussed the distinction between nesting and crossing with respect to facets in graphics, and he used `/` and `*` instead for nested and crossed expressions. A crossing structure displays all unique and possible combinations of crossed variables, including those not found in the data, in a graphical layout. However, tidy temporal data finds only the combinations that occur in the data and intentionally respects the structure of incomplete combinations.

# Data pipeline

This representation choice will prove suitable for almost all conceivable applications of temporal data. But the data abstraction cannot live alone with an additional toolchain in order to be able to acquire insight and knowledge about the data itself. A desirable toolbox chains transformation, visualisation, and modelling as a data science pipeline. Each component itself is also a self-contained pipeline that can be decomposed into many individual operations. This paper will focus on part of the general pipeline---data transformation, for example filtering observations, selecting and summarising variables. We shall see how this data abstraction breeds a consistent data process.

## Literature review

@@

## Time-based pipeline analysis

@@

An individual operation that performs on a specific task can be phrased as a verb. This verb is self-explanatory enough to express what the specific task is. We shall apply **dplyr** key verbs to time domain and expand its vocabulary a little more to handle time-related analysis problems.

>>>>

In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to different representations. A matrix is used to represent multivariate time series where each row represents observations measured at a time point and each column represents a series ("wide" form). This matrix representation requires homogeneity (that is, all the columns must be of same type.), time indices implicitly inferred as attributes or meta-information, series of same time length, and explicit missingness. By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm** [@R-plm]. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. less defaults to be made. [R]

@@

# Data semantics {#sec:semantics}

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represents time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" serves a similar purpose as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. A syntax is introduced to express a key consisting of nested and crossed factors. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database. Table \@ref(tab:tb-sub) shows a "tidy" temporal data example using a subset of tuberculosis cases estimated by @tb-data. It contains 12 observations and 5 variables arranged in a "long" tabular form. Each observation uniquely records the number of people, who are diagnosed tuberculosis for each gender at three selected countries in the years of 2011 and 2012, where column `year` is declared as the index variable, columns `country`, `continent` and `gender` as key. To be more specific, the key is passed via `id(country | continent, gender)` or `id(gender, country | continent)`: a geographic hierarchy of country nested within continent, crossed with a demographic factor. Detailed discussions about index and key specifications will be given in Section \@ref(sec:index) and \@ref(sec:key). Although column `count` is the only measure in this case, it is sufficiently flexible to hold other measured variables, for example, adding the corresponding population size (if known) in order to calibrate the count later.

Given the nature of temporal ordering, a temporal data set must be sorted by time index as a result. If a key is explicitly declared, the key will be sorted first depending on the ordering of factor levels and the ordering of variables, and followed by arranging time in ascending order. The tuberculosis example shows the key is sorted in alphabetic order, followed by ascending years by default, as is Table \@ref(tab:tb-sub). Since time-based operations, for example leading, lagging, or differencing series, require a vector in its time order, the arranged data format assures its validity for these operations without returning unexpected results. There are other possible arrangements in the key, but the baseline is the index (i.e. `year`) ascends within each key element. If any operations on the tidy temporal data results a deviation from this baseline, it starts issuing a warning to the users.

@@

```{r tb-sub}
```

(ref:tb-sub) A small subset of estimates of tuberculosis burden generated by World Health Organisation in 2011 and 2012, with 12 observations and 5 variables. The index refers to column `year`, the key to multiple columns: `contry` nested under `continent` crossed with `gender`, and the measured variable to column `count`.

## Time index and interval {#sec:index}

@@

For data indexed in regular time space, the time interval is obtained by computing the greatest common divisor based on the non-negative differences of a given time index, although the data may include implicit time gaps. This implies that all the units in the table must share a common interval. It is relatively easy to determine if the data is daily or sub-daily since most statistical packages have built-in time representations for date-times and dates, for example their corresponding classes `POSIXct` and `Date` in R. However these turn out inappropriate for weekly, monthly and quarterly data. For example, the smallest time unit of monthly data is months in a year, which is equispaced points in a year. However, using the first day of the month can range from 28 to 31 days, which is irregularly spaced. A genuine year-month representation should avoid associating days with its format. Displaying the year and month components suffice to signal its monthly aggregated values instead of observed values at a particular timestamp. This argument is also applicable to the creation of year-week and year-quarter. (coercion)

Since Table \@ref(tab:tb-sub) possesses complete observations for each category, the interval obtained from the greatest common factor is 1 year.

@@

## Key {#sec:key}

The "key" identifies units or subjects that are observed over time in a data table, which are typically known in advance by users. The absence of key only occurs to a univariate case where the key is implicit. Multiple units must associate with an explicit "key" to allow for identification. The key is not constrained to a single column, but is comprised of multiple columns. When involving multiple variables, one need to distinguish nesting and crossing data variables. "Nesting" refers to a scenario where a variable is nested within another when each category of the former variable co-occurs with only one category of the latter (for example, `country` nested within `continent`); on the other hand, "crossing" means that a variable is crossed with another when every category of one variable co-occurs with every category of the other (for example, `country` crosses with `gender`). Nesting is a special case of crossing. The former exhibits a hierarchical ordering with lower-level category (children) nested within higher-level category (parent) (more categories in lower level); whereas the ordering of the latter makes no difference. Whilst constructing tidy temporal data, users also need to specify the key that identifies the structure on the data. Rather than creating a separate object that contains the data structure, like what the **hts** does, we propose a syntax-based interface to directly deal with variables, so that it takes advantage of data semantics. The expression of a vertical bar (`|`) is used to indicate nesting variables, while a comma (`,`) for crossing variables. @wilkinson2006grammar discussed the distinction between nesting and crossing with respect to facets in graphics, and he used `/` and `*` instead for nested and crossed expressions. A crossing structure displays all unique and possible combinations of crossed variables, including those not found in the data, in a graphical layout. However, tidy temporal data finds only the combinations that occur in the data and intentionally respects the structure of incomplete combinations.

# Data pipeline {#sec:pipeline}

@@

@@

## Time-based pipeline

This "tidy" data representation choice will prove suitable for almost all conceivable applications of temporal data. But the data abstraction cannot live alone with an additional toolchain in order to be able to acquire insight and knowledge about the data itself. A desirable toolbox chains transformation, visualisation, and modelling as a data science pipeline. Each component itself is also a self-contained pipeline that can be decomposed into many individual operations. This paper will focus on part of the general pipeline---data transformation, for example filtering observations, selecting, and summarising variables. We shall see how this data abstraction breeds a consistent data process.

## Expressive verbs

An individual operation that performs on a specific task can be phrased as a verb. This verb is self-explanatory enough to express what it is supposed to do or fail. We shall adapt **dplyr** key verbs to time domain and expand its vocabulary a little more to handle time-related analysis problems.
